Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-07-03 manuscript/manuscript.tex
Beginning 2nd pass for document summary.

\section{NYC 311 Service Request Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year, covering issues such as street noise, illegal
parking, heat and hot water, abandoned vehicles, and unsanitary
conditions. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results.

Despite its success, the 311 system faces several challenges, such as
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, handling of personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usage and load for numerous
third-party users. These challenges are addressed by various agency
open data managers and the
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system. Data often needs to be exported and processed in
external programs to address these issues, as observed in the
examination of the large 311 dataset. Furthermore, the system must
adapt to changes such as agency name changes and the introduction of
new complaint types, which can complicate long-term data analysis.

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping city governance and community
engagement. This open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID-19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
improving responsiveness for city agencies like Parks and Recreation,
the Department of Transportation, and the Department of Sanitation,
and addressing homelessness through initiatives like the Homeless
Outreach and Mobile Engagement Street Action Teams (HOME-STAT). These
applications demonstrate the transformative potential of open data in
enhancing urban living and fostering a more responsive and transparent
city government.

\begin{comment}
\jy{FixME: this paragraph's location will be revisited later.}
We extracted 311 SR data for the period of CY 2022--2023, a dataset of
6.4 million SR records. This is a large dataset which enables 
observation of some rarely seen events, data inaccuracies, 
and logical inconsistencies. It provides enough data to 
examine trends and data outlier analysis. We queried the data 
directly from the NYC Open Data Portal and exported the 
data for analysis with custom written R programs.
\end{comment}

@@

To begin, we looked at a 10-year period (2014-2023) to
observe the data over a longer time-frame; see
Figure~\ref{fig:10-yr-monthly}. During that time-frame,
the number of 311 SRs grew by nearly 50\%. The summer months make
up the highest days, and the winter months show the lowest activity.

\begin{comment}
\begin{table}[tbp]
    \centering
    \caption{Top 5 and Bottom 5 Days for SRs}
    \label{tab:top-10-counts}
    \begin{tabular}{@{}lll lll@{}}
      \toprule
      \multicolumn{2}{c}{Top 5 Days} & \multicolumn{2}{c}{Bottom 5 Days} \\
      \cmidrule(r){1-2} \cmidrule(l){3-4}
      Date & Count & Date & Count\\
      \midrule
      2020-08-04 & 24,415 & 2014-12-25 & 2965\\
      2020-08-05 & 19,560 & 2015-12-25 & 3247\\
      2023-09-29 & 17,962 & 2014-04-20 & 3287\\
      2020-07-05 & 16,916 & 2015-12-26 & 3405\\
      2020-06-21 & 15,883 & 2016-11-06 & 3419\\
      \bottomrule
    \end{tabular}
\end{table}
\end{comment}

%Table~\ref{tab:top-10-counts} summarizes the top 5 and bottom 5 days
i%n counts of SRs.

The highest daily SR count, 24,415, occurred on
August 4, 2020, coinciding with the aftermath of Tropical Storm
Isaias, highlighting the system's critical role during crises.

% The second-highest count of 19,560 was the following day, 
%reflecting the storm's continued impact.

Conversely, the lowest daily SR count, 2,965, was on Christmas Day,
2014, with similar lows on other Christmas Days, indicating reduced
activity during major holidays.

% The data also shows fewer requests on weekends, such as 
%April 20, 2014 (Easter Sunday), and November 6, 2016 (a non-holiday Sunday).

@@

Figure~\ref{fig:SRcountbyAgency} display the SR counts by agency with
cumulative percentage. Two important fields in this analysis are the
responsible NYC Agency and the type of complaint. Identifying the 
responsible agency and type of complaint is essential 
for identifying the accountable party and troubleshooting discrepancies. 
This identification also assists in determining whether an issue is 
specific to an agency or indicative of a broader, systemic problem. 
Six agencies account for 90\% of the complaints. These are the New York
Police Department (NYPD), Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).

@@

The distribution of complaint types skews toward a few select
categories, accompanied by a long tail. Figure~\ref{fig:SR_complaints}
shows that the top 20 complaint types account for 70\% of the total
220 different types. Notably, these top complaints include several
variations of noise-related issues, which collectively represent 22\%
of all complaints. When consolidated, noise complaints emerge as the
most frequently occurring complaint type.

% Table~\ref{tab:noisecomplaints}
%provides a detailed breakdown of noise-related complaint types by
%count and responsible agency. The most common noise complaint is
%``Noise - Residential,'' with 675,502 complaints, accounting for 10.56\%
%of all service requests, and primarily handled by the NYPD. Other
%significant noise complaints include ``Noise - Street/Sidewalk"
(%4.70\%), ``Noise - Commercial'' (2.02\%), and ``Noise - Vehicle"
(%1.81\%), all managed by the NYPD. Additional categories such as
%"Noise - Helicopter'' (1.33\%) fall under the jurisdiction of the EDC,
%and ``Noise'' (1.57\%) is managed by the DEP. The prevalence of these
%complaints underscores the critical role of the NYPD and other
%agencies in addressing noise-related issues, reflecting the
%significant impact of urban noise on residents' quality of life. This
%data highlights the importance of efficient response and management
%strategies to mitigate noise disturbances in New York City.

>>>>

\section{NYC 311 Service Request Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results.

Despite its success, the 311 system faces several challenges:
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, excluding personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usagefor numerous
third-party users. These challenges are addressed by various agency
open data managers and the
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system.

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping city governance and community
engagement. This open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID-19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
improving responsiveness for city agencies, and addressing 
homelessness through initiatives like the Homeless
Outreach and Mobile Engagement Street Action Teams (HOME-STAT).

To begin, we looked at a 10-year period (2014-2023) to
observe the data over a longer time-frame; see
Figure~\ref{fig:10-yr-monthly}. During that time-frame,
the number of 311 SRs grew by nearly 50\%.

@@

@@

Figure~\ref{fig:SRcountbyAgency} displays SR counts by Agency with
cumulative percentage. Identifying the 
responsible agency and type of complaint is essential 
for troubleshooting discrepancies. 
This identification also assists in determining whether an issue is 
specific to an Agency or indicative of a broader, systemic problem. 
Six agencies account for 90\% of the complaints. These are the New York
Police Department (NYPD), Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).

@@

Figure~\ref{fig:SR_complaints} shows that the top 20 complaint 
types account for 70\% of the total 220 different types. Notably, 
these top complaints include several variations of noise-related 
issues, which collectively represent 22\% of all complaints, which
when consolidated is the most frequently occurring complaint type.
